A Higher-Order Approach to Explicit Contexts in LF

### 2015-03-28 paper/paper.tex
remove sketchy proofs for now
wordsmithing, conclusion

  \item The correctness of this encoding depends very strongly on its
    \emph{use}: this only truly encodes contexts if in every case an LF-variable
    is used for $X$.  Moreover, the various well-formedness constraints are a
    bit inelegant, as is the concept of mapping LF variables to De Bruijn indices.

@@

\section{Encoding Telescopes in Higher-Order Abstract Syntax}
More precisely, it seems that what we really need is an encoding of
\emph{telescopes}, in the sense of De Bruijn. A telescope is a
sequence of type assignments, where variables bound earlier may appear free
later; for instance:
\[
  \cdot,x_1:A_1, x_2:A_2(x_1), x_3:A_3(x_1,x_2), \dots, x_n:A_n(x_1,\dots,x_{n-1})
\]

@@

The next thing to do is to specify how each of these judgement(-codes) behaves
operationally (i.e. by substituting its outermost variable); we will do so by
extending the LF signature with a new algorithmic judgement
$\sploot{\dname{J}}{\dname{J}'}$ (``$\InputMode{\dname{J}}$ sploots into
$\OutputMode{\dname{J}'}$'') and making it evident on all inputs:

@@

Next, we will do something similar in order to show how each of these
judgement(-codes) behaves extensionally, by introducing another algorithmic
judgement $\refract{\dname{J}}{\dname{J}^=}$ (``$\InputMode{\dname{J}}$
refracts into $\OutputMode{\dname{J}^=}$''). Intuitively, this judgement
describes how what happens when you substitute two purportedly equal
expressions in for the judgement-code's outermost variable.

@@

@@

\subsection{Admissible Rules}

\theorem{The standard application rule for the cartesian product of a family of
sets is admissible:
\[
  \infer[\typi_E]{
    \mem{\ap(M;N)}{B(N)}
  }{
    \dlabel{\dname D}{
      \mem{M}{\thunk{\typi(A;B)}}
    } &
   \dlabel{\dname E}{
      \mem{N}{A}
    }
  }
\]
}
\begin{proof}
  By induction on the derivation $\dname{D}$, which must have the following canonical form:
  \[
     \infer[\dname{D}]{
       \mem{M}{\thunk{\typi(A;B)}}
     }{
       \dlabel{\dname D}{\eval{M}{\lam E}} &
       \infer{
         \canmem{\lam E}{\typi(A;B)}
       }{
         \infer{
           \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
         }{
           \dlabel{\dname D''}{
             \lfpi{x,x'}{
               \weaken{\cdot}{x}{x'}
               \to\sequent{\cdot}{\qmem{x'}{A}}
               \to\sequent{\cdot}{\qmem{\nobind{E(x)}}{\nobind{B(x)}}}
             }
           } &
           \dlabel{}{\dots}
         }
       }
     }
  \]

@@

\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
\[
  \infer[\tysg_{E_1}]{
    \mem{\fst(M)}{A}
  }{
    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
  }\quad
  \infer[\tysg_{E_2}]{
    \mem{\snd(M)}{B(\fst(M))}
  }{
    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
  }
\]
}
\begin{proof}

  The derivations of the elimination rules are constructed by induction on
  $\dname{D}$, which must be a canonical derivation of the following form:
  \[
    \infer[\dname{D}]{
      \mem{M}{\thunk{\tysg(A;B)}}
    }{
      \dlabel{\dname D_1}{
        \eval{M}{\pair(M_1,M_2)}
      } &
      \infer{
        \canmem{\pair(M_1,M_2)}{\tysg(A;B)}
      }{
        \infer{
          \mem{M_1}{A}
        }{
          \dlabel{\dname D_2}{\eval{M_1}{X}} &
          \dlabel{\dname D_3}{\eval{A}{A'}} &
          \dlabel{\dname D_4}{\canmem{X}{A'}}
        }&
        \infer[\dname D_5]{
          \mem{M_2}{B(M_1)}
        }{
          \dlabel{\dname D_6}{\eval{M_2}{Y}} &
          \dlabel{\dname D_7}{\eval{B(M_1)}{C}} &
          \dlabel{\dname D_8}{\canmem{Y}{C}}
        }
      }
    }
  \]

  The two elimination rules are derived as follows:
  \medskip

  \framebox{
    \begin{minipage}{0.9\textwidth}
      \begin{gather*}
        \infer[\tysg_{E_1}]{
          \mem{\fst(M)}{A}
        }{
          \infer{
            \eval{\fst(M)}{X}
          }{
            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
            \dlabel{\dname D_2}{\eval{M_1}{X}}
          } &
          \dlabel{\dname D_3}{\eval{A}{A'}} &
          \dlabel{\dname D_4}{\canmem{X}{A'}}
        }\\[12pt]
        \infer[\tysg_{E_2}]{
          \mem{\snd(M)}{B(\fst(M))}
        }{
          \infer{
            \eval{\snd(M)}{Y}
          }{
            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
            \dlabel{\dname D_6}{\eval{M_2}{Y}}
          }&
          \dlabel{\dname D_7}{\eval{B(\fst(M))}{C}} & % TODO: A little fishy
          \dlabel{\dname D_8}{\canmem{Y}{C}}
        }
      \end{gather*}
    \end{minipage}
  }
\end{proof}

>>>>

  \item The adequacy of this encoding depends very strongly on its
    \emph{use}: this only truly encodes contexts if in every case an LF-variable
    is used for $X$.  Moreover, the various well-formedness constraints are a
    bit inelegant, as is the concept of mapping LF variables to De Bruijn indices.

@@

\section{Encoding Telescopes in Higher-Order Abstract Syntax}
More precisely, it seems that what we really need is an encoding of
\emph{telescopes}, in the sense of De
Bruijn\cite{Bruijn91telescopicmappings}. A telescope is a sequence of
type assignments, where variables bound earlier may appear free later;
for instance:
\[
  \cdot,x_1:A_1, x_2:A_2(x_1), x_3:A_3(x_1,x_2), \dots, x_n:A_n(x_1,\dots,x_{n-1})
\]

@@

The next thing to do is to specify how each of these judgement(-codes) behaves
operationally (i.e. by substituting its outermost variable); we will do so by
extending the LF signature with a new algorithmic judgement
$\sploot{\dname{J}}{\dname{J}'}$ and making it evident on all inputs:

@@

Next, we will do something similar in order to show how each of these
judgement(-codes) behaves extensionally, by introducing another algorithmic
judgement $\refract{\dname{J}}{\dname{J}^=}$. Intuitively, this judgement
describes how what happens when you substitute two purportedly equal
expressions in for the judgement-code's outermost variable.

@@

The judgement $\sploot{\dname{J}}{\dname{J}'}$ may be considered as
defining the behavior of $\dname{J}$ with one hole punched in it;
$[x]\;\dname{J}(x)$ could be called the ``one-hole application of
$\dname{J}$''. Likewise, $\refract{\dname{J}}{\dname{J}^=}$ may be
thought of as defining the behavior $\dname{J}$ with two hole punched
in it, and $[x][y]\; \dname{J}^=(x,y)$ could be called the ``two-hole
application of $\dname{J}$''.

@@

\subsection{Validity of Elimination Rules}

Here, we demonstrate the validity of the application rule for $\typi$
sets; the validity of further elimination rules is left for future
work.

\begin{theorem}
The standard application rule for the cartesian product of a family of
sets is admissible:
\[
  \infer[\typi_E]{
    \mem{\ap(M;N)}{B(N)}
  }{
    \dlabel{\dname D}{
      \mem{M}{\thunk{\typi(A;B)}}
    } &
   \dlabel{\dname E}{
      \mem{N}{A}
    }
  }
\]
\end{theorem}

\begin{proof}
  By induction on the derivation $\dname{D}$, which must have the following canonical form:
  \[
     \infer[\dname{D}]{
       \mem{M}{\thunk{\typi(A;B)}}
     }{
       \dlabel{\dname D}{\eval{M}{\lam E}} &
       \infer{
         \canmem{\lam E}{\typi(A;B)}
       }{
         \infer{
           \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
         }{
           \dlabel{\dname D''}{
             \lfpi{x,x'}{
               \weaken{\cdot}{x}{x'}
               \to\sequent{\cdot}{\qmem{x'}{A}}
               \to\sequent{\cdot}{\qmem{\nobind{E(x)}}{\nobind{B(x)}}}
             }
           } &
           \dlabel{}{\dots}
         }
       }
     }
  \]

@@

\section{Conclusion}

A rich, higher-order encoding for telescopes addresses several issues
which come up when encoding Martin-L\"of's extensional type theory
into the LF.

First, functionality premises no longer need be directly included in
every rule which uses hypothetico-general judgement. These are
factored out into a separate judgement which has a compositional
definition (the functional sequent judgement). Because of this, the
rules in the LF will look very much like how they did in
Mpartin-L\"of's informal presentation.

Secondly, the core judgements of type theory (i.e.\ typehood, equality
of types, membership, and equality of members) are defined first in
their atomic/categorical form, and then separately they are explained
with respect to a telescope, both directly and extensionally. Unlike
in Martin-L\"of's presentation, the sequent judgement is, then, not
defined separately at each categorical judgement, but is rather
explained once and for all in terms of the novel
$\sploot{\dname{J}}{\dname{J}'}$ and
$\refract{\dname{J}}{\dname{J}^=}$ judgements.
